Equation-free analysis of a dynamically evolving multigraph

### 2016-02-09 report/multigraph_manuscript.tex
moar manuscript moar referencez

Over the past decades, myriad systems have been formulated as complex evolving networks, from highway traffic \cite{joubert_large-scale_2010} to brain connectivity \cite{hermundstad_learning_2011}. When modeling such systems, dynamics are typically defined at a very fine scale, specifying individual node and edge interactions; explicit closed equations governing macroscopic properties are often unavailable \cite{durrett_graph_2012,joubert_large-scale_2010,roche_agent-based_2011,swaminathan_modeling_1998}. These interactions are often complicated functions dependent on multiple system parameters, which, when coupled with the large network sizes inherent in many interesting problems, makes numerical investigation computationally prohibitive. However, the lack of explicit equations governing macroscopic property evolution precludes the use of traditional numerical techniques such as optimization and fixed-point finding that could offer valuable insight into the network's behavior, leaving little alternative but to work with the full network system. Faced with this dilemma, investigators must either restrict their attention to a smaller, heuristically-motivated parameter space \cite{hodgkin_quantitative_1952} or simplify the model, potentially removing important features \cite{brown_variability_1999}. \par

Equation-free modeling offers a solution to these obstacles\ignore{circumvents these obstacles}, allowing one to investigate the complex network at a macroscopic level while retaining full system details. Underlying this method is the assumption that, although we cannot analytically derive equations governing network evolution, such equations do exist. Furthermore, these implicit equations are assumed to be functions of only a few dominating variables. Thus, the state of the complete network is in fact, by this assumption, well described by these select variables. This may seem an onerous restriction, yet it is exactly the behavior witnessed across network types: despite the complexity of the full system, certain network properties evolve smoothly in time, while the evolution of other, secondary properties, is often correlated to that of a few primary variables. Once these significant variables are uncovered, we can combine short intervals of full system simulation with operators that map the full system description to and from their representative coarse variables to perform previously-unfeasible system-level analysis (see Section \ref{sec:ef} for further details). \par

@@

We begin in Section \ref{sec:m} with a description of the model. Section \ref{sec:ef} provides further details of the equation-free method, specifies how it was applied to our system, and presents subsequent results. Section \ref{sec:dr} summarizes the operation of PCA and DMAPS, and assesses their effectiveness in analyzing hidden, low-dimensional structure in the model.

@@

We study the edge-conservative preferential attachment model, a detailed mathematical analysis of which can be found in \cite{rath_time_2009} and \cite{rath_multigraph_2012}. The system evolves in discrete steps $t = 0,1,\ldots t_f$, and we denote the $n$-vertex graph at each point by $G_n(t)$. The initial state, $G_n(0)$, is an arbitrary distribution of $m$ edges among the $n$ vertices. No restrictions are placed on this initial distribution: multiple edges and loops are permitted.\ignore{an Erd\H{o}s-R\'{e}nyi random graph on $n$ vertices and $m$ edges. Here, however, $m$ may be larger than the total number of edges in a completely-connected $n$-vertex graph so we cannot assign a probability to each edges existence, instead placing $m$ edges at random in the graph.} The system is then advanced step-by-step based on the following procedure:

@@

Evolving the system in this manner, the degree sequence approaches a stationary distribution over $O(n^3)$ steps. As explained in \cite{rath_time_2009}, this distribution is dependent only on system parameters $\rho = \frac{2*m}{n}$ and $\kappa$. Fig. (\ref{fig:dse}) illustrates the evolution of the degree sequence of two networks with different initial configurations but equivalent values of $\rho$ and $\kappa$; thus, we observe they approach an identical stationary state.

@@

However, it is crucial to test whether the evolution of other properties of the graph can be correlated to degree sequence. If not, our current description is incomplete and there exist other  important coarse variables that must be accounted for in our macroscopic system description. To assess this, we constructed graphs with identical degree sequences but varying betweenness-centrality distributions and clustering coefficients, and observed the evolution of these properties under the dynamics prescribed in Section \ref{sec:m}. Fig. (\ref{fig:sv}) shows that, within a short time, both variables quickly reach similar distributions despite the varied initial conditions. This suggests that the degree sequence is indeed an adequate coarse variable for the system. \par

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=\textwidth]{cpi-diagram}
  \caption{Schematic of our restriction ($R$) and lifting ($L$)
    procedures. The graph's sorted degree sequence is calculated
    ($R_1$) and then fit to a third-order polynomial ($R_2$). This
    maps graphs ($G$) to vectors of four coefficients
    ($\mathbf{c}$). To lift, the polynomial is used to generate a
    degree sequence ($L_1$), which is then used to construct a
    consistent graph ($L_2$). This process maps vectors of
    four coefficients ($\mathbf{c}$) to graphs
    ($G$). $R = R_2 \circ R_1 $ is our restriction and $L = L_2 \circ
    L_1$ is our lifting. \label{fig:cpi-diagram}}
\end{figure}

@@

The restriction operator translates the full network into its equivalent coarse variables. Here, this involves mapping a multigraph to its sorted degree sequence, a simple calculation. However, we can continue to reduce the dimensionality of our coarse system by keeping not the length-$n$ degree sequence, but a polynomial fitting of it. To do so, we first sort the sequence to achieve a smooth, monotonically increasing dataset, then fit the result to some function. Here, a sixth-degree polynomial was observed to closely approximate the sequence throughout time; thus, our coarse description consisted of the seven polynomial coefficients specifying a particular sorted degree sequence, as shown in Fig. (\ref{fig:cpi-diagram}). The restriction operator therefore maps multigraphs to length-seven coefficient vectors. This procedure is represented by the blue arrows of Fig. (\ref{fig:cpi-diagram}).

@@

Aside from the computational savings of model simulations offered by CPI, the equation free framework also permits the calculation of system steady states through matrix-free Newton-GMRES algorithms. Referring back to the CPI procedure outlined in Sec. (\ref{sec:cpi}), we can define an operator $\Phi: c \rightarrow c$ projecting coarse variables at $t$ to their values at $t + \delta t$: $c(t+\delta t) =\Phi(c(t))$. A system fixed point could then be located by employing Newton's method to solve the equation

\begin{align}
\label{eq:f}
F(c) = \Phi(c) - c = 0
\end{align}

However, this requires the repeated solution of $DF(c^{(k)}) \delta c^{(k+1)} = -F(c^{(j)})$, in which the system Jacobian, $DF$ is unavailable to us. Thankfully, we may circumvent this obstacle by estimating the directional derivatives $DF(c) \cdot \delta c$ via a difference approximation of the form

\begin{align}
  DF(c) \cdot \delta c \approx \frac{\| \delta c \| F(c + h \| c \| \frac{\delta c}{\| \delta c \|}) - F(c)}{h \| c \|}
\end{align}

for nonzero $c$, which in turn is evaluated through calls to $F$ as defined in Eq. (\ref{eq:f})

This is precisely the approach of the Newton-GMRES method, in which the solution to a linear system is searched for in expanding Krylov subspaces \cite{kelley_solving_2003}. Applying this algorithm in conjunction with the $\Phi$ operator defined in Sec. (\ref{sec:cpi}) allowed us to locate the stationary distribution without simply running and observing the full system for long times, as would otherwise be required. Results are shown in Fig. (\ref{fig:newton-results}).

@@

When applying machine learning techniques to a dataset, it is
necessary to define a distance (or similarity) between each pair of
data point. If these points are a set of vectors in $\mathbb{R}^n$ one
has a number of metrics to choose from, the Euclidean distance being a
common first option. Unfortunately, when individual points are not
vectors but networks, the definition of a computationally feasible
metric becomes far more challenging. Examples such as the maximal
common subgraph and edit distances, defined in \cite{bunke_graph_1998}
and \cite{gao_survey_2010} do define metrics on the space of graphs,
but their computation is $NP\textnormal{-}hard$. Other computationally
feasible approaches include comparing distributions of random
walks on each graph \cite{vishwanathan_graph_2010}, calculating the so-called
$n$-tangle density \cite{gallos_revealing_2014}, or calculting the edit distance
with an approximation algorithm \cite{riesen_approximate_2009,zeng_comparing_2009}.  \\

The strategy used in the following methods, detailed in
\cite{rajendran_analysis_2013} and \cite{xiao_structure-based_2008}
enumerates the number of times a certain set of motifs (or subgraphs)
appears in each network in the dataset. This maps each network to a
vector in $\mathbb{R}^n$, and the Euclidean distance is subsequently
used to calculate the similarity of two graphs. Due to computational
restrictions, we chose to count the number of three- and four-vertex
subgraphs that were contained in each network. As there are eight such
motifs, shown in Fig. (\ref{fig:motifs}), this process $\gamma$ maps
each graph to an eight-dimensional vector:
$\gamma : G \rightarrow \mathbb{R}^8$. Additionally, we applied this
operateration to a simplified version of each graph, wherein any
multiple edges were reduced to single edges.

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \includegraphics[width=\textwidth]{motifs}
  \caption{List of the subgraphs used to embed each network. The
    number of times each appeared in the simplified input graph was
    calculated, effectively mapping graphs to $\mathbb{R}^8$. \label{fig:motifs}}
\end{figure}

@@

PCA is used to linearly project data into subspaces that capture the
directions along which the data varies most
\cite{jolliffe_principal_2014}. It has found wide application, but
can suffer from its restriction to linear embeddings. Indeed, many
datasets will not lie along some hyperplane embedded in
$\mathbb{R}^n$, but rather will form a nonlinear manifold throughout
the space. \\

Theoretical results in \cite{rath_time_2009} state that, for a given network
size $n$, the final stationary state depends only on the number of
edges present $m$ and the model parameter $\kappa$. To assess both the
validity of our graph embedding technique and the usefullness of the
PCA in the context of complex networks, we generated a dataset of
stationary states over a range of $m$ and $\kappa$ values by directly
running the model over $O(n^3)$ steps ($N$ values of each parameter
were taken, for a total number of $N^2$ networks). Each resulting
graph $G(m_i, \kappa_j) = G_{ij}$ was then embedded into
$\mathbb{R}^8$ by counting the number of times each of the subgraphs
shown in Fig. (\ref{fig:motifs}) appeared in the network. Thus
$\gamma(G_{ij}) = v_{ij} \in \mathbb{R}^8$. We then proceeded to
perform a principal component analysis on this collection of vectors
${v_{ij}}_{i,j=1}^N$. Surprisingly, this linear technique succeeded in
uncovering a two-dimensional embedding of the dataset corresponding to
the two underlying parameters $m$ and $\kappa$, as shown in
Fig. (\ref{fig:pca}). This suggests that, given some final network
state $G_f(m, \kappa)$, by projecting its embedding onto these first two
principal components one could obtain a reasonable approximation of
the hidden parameter $\kappa$.

@@

Unlike PCA, DMAPS uncovers underlying nonlinear manifolds hidden in data. This is achieved by solving for the discrete equivalent of the eigenfunctions and eigenvalues of the Laplacian operator over the manifold, which amounts to calculating leading eigenvector/eigenvalue pairs of a symmetric Markov matrix $W$. See \cite{coifman_diffusion_2006,nadler_diffusion_2006} for details. \\

First we applied DMAPS to the dataset described in
Sec. (\ref{sec:pca}). Given the relative success of PCA in this
setting, one would expect DMAPS to also uncover the two-dimensional
embedding corresponding to different values of $m$ and $\kappa$. As
Fig. (\ref{fig:dmaps-rk}) shows, this is indeed the
case. Additionally, we were interested in embeddings of different
model trajectories. This dataset was generated by sampling two
different model simulations at even intervals as they evolved ($N$
points were sampled from each trajectory). The parameter values $m$
and $\kappa$ were held constant, but one graph was initialiazed as an
Erd\H{o}s-R\'{e}nyi random graph (Fig. (\ref{fig:erdos-init})), while
the other was initialized as a lopsided graph
(Fig. (\ref{fig:lopsided-init})). Letting $G_e(t)$ refer to the
Erd\H{o}s-R\'{e}nyi-initialized system at step $t$, and $G_l(t)$ the
lopsided-initialized system, the embedding $\gamma$ was used to create
points $\gamma(G_e(t) = v_e(t) \in \mathbb{R}^8$ and similarly
$\gamma(G_l(t) = v_l(t) \in \mathbb{R}^8$. \\

DMAPS was then applied to this set of $2N$ points, the results of
which are shown in Fig. (\ref{fig:dmaps-results}). At first, the two
points are mapped to distant regions in $\mathbb{R}^3$ due to their
different initial conditions. They also evolve along different
coordinates of the embedding, showing they traverse two distinct
trajectories on their approach to the stationary state. The particular
eigenvector each trajectory travels along is also significant: as
$G_e(t)$ was mapped to $\Phi_1$ and $G_l(t)$ to $\Phi_2$, we can infer
that the dynamics of the Erd\H{o}s-R\'{e}nyi-initialized system occur
more slowly than those of the lopsided-initialized system. Indeed,
referring back to Fig. (\ref{fig:dse}), we see that the lopsided
network descends more rapidly to the stationary state. Eventually,
their embeddings meet as they arrive at this final, shared state. Thus
DMAPS has proven useful in elucidating both geometric and dynamic
features of the system as shown in Figs. (\ref{fig:dmaps-rk}) and (\ref{fig:dmaps-results}). \\

>>>>

Over the past decades, myriad systems have been formulated as complex
evolving networks, from highway traffic
\cite{joubert_large-scale_2010} to brain connectivity
\cite{hermundstad_learning_2011}. When modeling such systems, dynamics
are typically defined at a very fine scale, specifying individual node
and edge interactions; explicit closed equations governing macroscopic
properties are often unavailable
\cite{durrett_graph_2012,joubert_large-scale_2010,roche_agent-based_2011,swaminathan_modeling_1998}. These
interactions are often complicated functions dependent on multiple
system parameters, which, when coupled with the large network sizes
inherent in many interesting problems, makes numerical investigation
computationally prohibitive. Additionally, the lack of such explicitly
macroscopic equations precludes the use of traditional numerical
techniques such as optimization and fixed-point finding that could
offer valuable insight into the network's behavior, leaving little
alternative but to work with the full network system. Faced with this
dilemma, investigators must either restrict their attention to a
smaller, heuristically-motivated parameter space
\cite{hodgkin_quantitative_1952} or simplify the model, potentially
removing important features \cite{brown_variability_1999}. \par

Equation-free modeling offers a solution to these obstacles, allowing
one to investigate the complex network at a macroscopic level while
retaining full system details \cite{kevrekidis_equation-free:_2004,gear_equation-free_2003}. Underlying this method is the
assumption that, although we cannot analytically derive equations
governing network evolution, such equations do exist. Furthermore,
these implicit equations are assumed to be functions of only a few
dominating variables. Thus, the state of the complete network is in
fact, by this assumption, well described by these select
quantities. This may seem an onerous restriction, yet it is exactly
the behavior witnessed across network types: despite the complexity of
the full system, certain network properties evolve smoothly in time,
while the evolution of other, secondary properties, is often
correlated to that of a few primary variables
\cite{bold_equation-free_2014,rajendran_coarse_2011,siettos_equation-free_2011}. Once
these significant variables are uncovered, we can combine short
intervals of full system simulation with operators that map the full
system description to and from their representative coarse variables
to perform previously-unfeasible system-level analysis (see Section
\ref{sec:ef} for further details). \par

@@

We begin in Section \ref{sec:m} with a description of the
model. Section \ref{sec:ef} provides further details of the
equation-free method, specifies how it was applied to our system, and
presents subsequent results. Section \ref{sec:dr} summarizes the
operation of PCA and DMAPS, and assesses their effectiveness in
analyzing hidden, low-dimensional structure in this network system.

@@

We study the edge-conservative preferential attachment model, a
detailed mathematical analysis of which can be found in
\cite{rath_time_2012} and \cite{rath_multigraph_2012}. The system
evolves in discrete steps $t = 0,1,\ldots t_f$, and we denote the
$n$-vertex graph at each point by $G_n(t)$. The initial state,
$G_n(0)$, is an arbitrary distribution of $m$ edges among the $n$
vertices. No restrictions are placed on this initial distribution:
multiple edges and loops are permitted.\ignore{an Erd\H{o}s-R\'{e}nyi
  random graph on $n$ vertices and $m$ edges. Here, however, $m$ may
  be larger than the total number of edges in a completely-connected
  $n$-vertex graph so we cannot assign a probability to each edges
  existence, instead placing $m$ edges at random in the graph.} The
system is then advanced step-by-step based on the following procedure:

@@

Evolving the system in this manner, the degree sequence approaches a
stationary distribution over $O(n^3)$ steps. As explained in
\cite{rath_time_2012}, this distribution is dependent only on system
parameters $\rho = \frac{2*m}{n}$ and $\kappa$. Fig. (\ref{fig:dse})
illustrates the evolution of the degree sequence of two networks with
different initial configurations but equivalent values of $\rho$ and
$\kappa$; as expected, we observe they approach an identical stationary state.

@@

However, it is crucial to test whether the evolution of other
properties of the graph can be correlated to degree sequence. If not,
our current description is incomplete and there exist other important
coarse variables that must be accounted for in our macroscopic system
description. To assess this, we constructed graphs with identical
degree sequences but varying triangle counts and observed the
evolution of this property under the dynamics prescribed in Section
\ref{sec:m}. Fig. (\ref{fig:sv}) shows that, within a short time, the
triangle counts are drawn to a shared slow manifold, despite the varied
initial conditions. This suggests that the degree sequence is indeed
an adequate coarse variable for the system. \par

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=\textwidth]{cpi-diagram}
  \caption{Schematic of our restriction ($R$) and lifting ($L$)
    procedures. The graph's sorted degree sequence is calculated
    ($R_1$) and then fit to a third-order polynomial ($R_2$). This
    maps graphs ($G$) to vectors of four coefficients
    ($\mathbf{c}$). To lift, the polynomial is used to generate a
    degree sequence ($L_1$), which is then used to construct a
    consistent graph ($L_2$). This process maps vectors of
    four coefficients ($\mathbf{c}$) to graphs
    ($G$). $\mathbf{R} = R_2 \circ R_1 $ is our restriction and $\mathbf{L} = L_2 \circ
    L_1$ is our lifting. \label{fig:cpi-diagram}}
\end{figure}

@@

The restriction operator translates the full network into its
equivalent coarse variables. Here, this involves mapping a multigraph
to its sorted degree sequence, a simple calculation. However, we can
continue to reduce the dimensionality of our coarse system by keeping
not the length-$n$ degree sequence, but a polynomial fitting of it. To
do so, we first sort the sequence to achieve a smooth, monotonically
increasing dataset, then fit the result to some function. Here, a
third-order polynomial was observed to closely approximate the
sequence throughout time; thus, our coarse description consisted of
the four polynomial coefficients specifying a particular sorted degree
sequence, as shown in Fig. (\ref{fig:cpi-diagram}). The restriction
operator therefore maps multigraphs to length-four coefficient
vectors. This procedure is represented by the blue arrows of
Fig. (\ref{fig:cpi-diagram}).

@@

Aside from the computational savings of model simulations offered by
CPI, the equation free framework also permits the calculation of
system steady states through matrix-free Newton-GMRES
algorithms. Referring back to the CPI procedure outlined in
Sec. (\ref{sec:cpi}), we can define an operator
$\Theta: d \rightarrow d$ projecting coarse variables at $t$ to their
values at $t + \delta t$: $d(t+\delta t) =\Theta(d(t))$. Note that in
this section, we take the sorted degree sequence as our coarse
variable. A system fixed point could then be located by employing
Newton's method to solve the equation

\begin{align}
\label{eq:f}
F(d) = \Theta(d) - d = 0
\end{align}

However, this requires the repeated solution of
$DF(d^{(k)}) \delta d^{(k+1)} = -F(d^{(j)})$, in which the system
Jacobian, $DF$ is unavailable to us. Thankfully, we may circumvent
this obstacle by estimating the directional derivatives
$DF(d) \d dot \delta d$ via a difference approximation of the form

\begin{align}
  DF(d) \ddot \delta d \approx \frac{\| \delta d \| F(d + h \| d \| \frac{\delta d}{\| \delta d \|}) - F(d)}{h \| d \|}
\end{align}

for nonzero $d$, which in turn is evaluated through calls to $F$ as defined in Eq. (\ref{eq:f})

This is precisely the approach of the Newton-GMRES method, in which the solution to a linear system is searched for in expanding Krylov subspaces \cite{kelley_solving_2003}. Applying this algorithm in conjunction with the $\Theta$ operator defined in Sec. (\ref{sec:cpi}) allowed us to locate the stationary distribution without simply running and observing the full system for long times, as would otherwise be required. Results are shown in Fig. (\ref{fig:newton-results}).

@@

When applying dimensionality reduction techniques to a dataset, it is
necessary to define a distance (or similarity) between each pair of
data point. If these points are a set of vectors in $\mathbb{R}^n$ one
has a number of metrics to choose from, the Euclidean distance being a
common first option. Unfortunately, when individual points are not
vectors but networks, the definition of a computationally feasible
metric becomes far more challenging. Examples such as the maximal
common subgraph and edit distances, defined in \cite{bunke_graph_1998}
and \cite{gao_survey_2010} do define metrics on the space of graphs,
but their computation is $NP\textnormal{-}hard$. Other computationally
feasible approaches include comparing distributions of random
walks on each graph \cite{vishwanathan_graph_2010}, calculating the so-called
$n$-tangle density \cite{gallos_revealing_2014}, or calculating the edit distance
with an approximation algorithm \cite{riesen_approximate_2009,zeng_comparing_2009}.  \\

The strategy used in the following methods, detailed in
\cite{rajendran_analysis_2013} and \cite{xiao_structure-based_2008}
enumerates the number of times a certain set of motifs (or subgraphs)
appears in each network in the dataset. This maps each network to a
vector in $\mathbb{R}^n$, and the Euclidean distance is subsequently
used to calculate the similarity of two graphs. Due to computational
restrictions, we chose to count the number of three- and four-vertex
subgraphs that were contained in each network. As there are eight such
motifs, shown in Fig. (\ref{fig:motifs}), this process $\gamma$ maps
each graph to an eight-dimensional vector:
$\gamma : G \rightarrow \mathbb{R}^8$. Additionally, we applied this
operation to a simplified version of each graph, wherein any
multiple edges were reduced to single edges.

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \includegraphics[width=\textwidth]{motifs}
  \caption{List of the subgraphs used to embed each network. The
    number of times each appeared in the simplified input graph was
    calculated, mapping input graphs to $\mathbb{R}^8$. \label{fig:motifs}}
\end{figure}

@@

PCA is used to linearly project data into subspaces that capture the
directions along which the data varies most
\cite{jolliffe_principal_2014}. Given some matrix
$X \in \mathbb{R}^{n \times m}$ in which each of the $m$ column
vectors represent a different collection of the $n$ system variables,
PCA computes a reduced set of $k$ ``principal components''
$z_i \in \mathbb{R}^n$ with $k < n$ that constitute an optimal basis
for the data in the sense that, among linear, $k$-dimensional embeddings, $W = Z^TX$
captures the maximum possible variance in the dataset. \\

This method has found wide application, but can suffer from its
inability to uncover nonlinear relationships among the input
variables. Indeed, many datasets will not lie along some hyperplane
embedded in $\mathbb{R}^n$, but rather will form a nonlinear manifold
throughout the space. \\

Theoretical results in \cite{rath_time_2012} state that, for a given network
size $n$, the final stationary state depends only on the number of
edges present $m$ and the model parameter $\kappa$. To assess both the
validity of our graph embedding technique and the usefulness of the
PCA in the context of complex networks, we generated a dataset of
stationary states over a range of $m$ and $\kappa$ values by directly
running the model over $O(n^3)$ steps ($N$ values of each parameter
were taken, for a total of $N^2$ networks). Each resulting
graph $G(m_i, \kappa_j) = G_{ij}$ was then embedded into
$\mathbb{R}^8$ by counting the number of times each of the subgraphs
shown in Fig. (\ref{fig:motifs}) appeared in the network. Thus
$\gamma(G_{ij}) = v_{ij} \in \mathbb{R}^8$. We then proceeded to
perform a principal component analysis on this collection of vectors
$\{v_{ij}\}_{i,j=1}^N$. Surprisingly, this linear technique succeeded in
uncovering a two-dimensional embedding of the dataset corresponding to
the two underlying parameters $m$ and $\kappa$, as shown in
Fig. (\ref{fig:pca}). This suggests that, given some final network
state $G(m, \kappa)$, by projecting its embedding onto these first two
principal components one could obtain a reasonable approximation of
the hidden parameter $\kappa$.

@@

Unlike PCA, DMAPS uncovers nonlinear manifolds hidden in data. This is
achieved by solving for the discrete equivalent of the eigenfunctions
and eigenvalues of the Laplace-Beltrami operator over the manifold,
which amounts to calculating leading eigenvector/eigenvalue pairs of a
Markov matrix $A$ describing a diffusion process on the dataset. As the
eigenfunctions of the Laplace-Beltrami operator provide
parameterizations of the underlying domain, the output eigenvectors $\Phi_i$
from DMAPS similarly reveal and parameterize any hidden nonlinear
structure in the input dataset. The $k$-dimensional DMAP of point
$x_i$ is given by

\begin{align*}
  \Psi(x_i; t) = \begin{bmatrix} \lambda_1^t \Phi_1(i) \\ \lambda_2^t
    \Phi_2(i) \\ \vdots \\
    \lambda_k^t \Phi_k(i) \end{bmatrix}
\end{align*}

where $\lambda_i$ is the $i^{th}$ eigenvalue of the Markov matrix $A$,
$\Phi_i(j)$ the $j^{th}$ entry of eigenvector $i$, and the parameter
$t$ allows one to probe multiscale features in the dataset. See
\cite{coifman_diffusion_2006,nadler_diffusion_2006} for
further details. \\

First we applied DMAPS to the dataset described in
Sec. (\ref{sec:pca}). Given the relative success of PCA in this
setting, one would expect DMAPS to also uncover the two-dimensional
embedding corresponding to different values of $m$ and $\kappa$. As
Fig. (\ref{fig:dmaps-rk}) shows, this is indeed the case: using
$\Phi_1$ and $\Phi_4$ to embed the graphs produces a two-dimensional
surface along which both $\rho$ and $\kappa$ vary independently. \\

Additionally, we were interested in embeddings of different
model trajectories. This dataset was generated by sampling two
different model simulations at even intervals as they evolved ($N$
points were sampled from each trajectory). The parameter values $m$
and $\kappa$ were held constant, but one graph was initialized as an
Erd\H{o}s-R\'{e}nyi random graph (Fig. (\ref{fig:erdos-init})), while
the other was initialized as a lopsided graph
(Fig. (\ref{fig:lopsided-init})). Letting $G_e(t)$ refer to the
Erd\H{o}s-R\'{e}nyi-initialized system at step $t$, and $G_l(t)$ the
lopsided-initialized system, the embedding $\gamma$ was used to create
points $\gamma(G_e(t)) = v_e(t) \in \mathbb{R}^8$ and similarly
$\gamma(G_l(t)) = v_l(t) \in \mathbb{R}^8$. \\

DMAPS was then applied to this set of $2N$ points, and the
three-dimensional embedding using $\Phi_1$, $\Phi_2$ and $\Phi_3$ is
shown in Fig. (\ref{fig:dmaps-results}). At first, the two points are
mapped to distant regions in $\mathbb{R}^3$ due to their different
initial conditions. They evolve along different coordinates of
the embedding, indicating they traverse two distinct trajectories on
their approach to the stationary state. The particular eigenvector
each trajectory travels along is also significant: as $G_e(t)$ was
mapped to $\Phi_1$ and $G_l(t)$ to $\Phi_2$, we can infer that the
dynamics of the Erd\H{o}s-R\'{e}nyi-initialized system occur more
slowly than those of the lopsided-initialized system. Indeed,
referring back to Fig. (\ref{fig:dse}), we see that the lopsided
network descends more rapidly to the stationary state. Eventually,
their embeddings meet as they arrive at this final, shared state. Thus
DMAPS has proven useful in elucidating both geometric and dynamic
features of the system as shown in Figs. (\ref{fig:dmaps-rk}) and (\ref{fig:dmaps-results}). \\
